---
title: "Performing survival assesments for hydropower and pump compliance: Considerations for test choice, estimation and ethical implications"
author: "Josh Norman"
date: "2025-07-08"
output: html_document
---

# 1. Required libraries

```{r libraries, echo=FALSE}
#library(here)
library(tidyverse)
library(binom)
library(scales)
library(plotly)
library(survey) 
library(htmltools)
library(pwr)
library(TOSTER)
library(cowplot)
library(gsDesign)
#library(gganimate)
#library(ggpubr)
#library(gt)
#library(kableExtra)

#Source functions for calculations, plots etc
source("./functions.R")

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = getwd())
```

# 2. High survival estimation

For high survival scenarios rely on estimation instead of hypothesis-based testing.

This analysis aims to produce a viable Fixed-Width Sequential Confidence Interval Testing method for designing live fish trials with defined points to accept or reject null hypothesis and continue testing.

Fixed-precision Sequential Confidence Interval Testing

Fixed-width sequential confidence interval testing is a statistical method used to estimate a parameter (e.g., mean, proportion) with a predetermined precision while dynamically adjusting the sample size based on accumulating data. Unlike fixed-sample designs, this approach continuously monitors the width of the confidence interval (CI) during data collection, stopping only once the CI narrows to a pre-specified fixed width, ensuring the estimate meets the desired margin of error. Sequential testing is particularly useful when data collection is costly or time-consuming, as it often requires fewer observations than fixed-sample methods while maintaining statistical validity.

We deviate from this principle by fixing our precision level as apposed to the width of the CI.

Ensure we calculate our confidence intervals based on the survival probability, NOT, the mortality probability. This may be a source of error for the previous differences in calculations. 

At each sampling point (e.g., every n = 100), calculate Wilson confidence intervals for observed mortality number. This will inherently deviate from fixed-width CI as the CI width is determined by observed mortality and sample size.

Decision rules to estimate probability with a given precision level (differs from hypothesis based testing):

-   Desired threshold for survival ≥ 98%

-   Not fish-friendly : If upper 95% CI \< 98%

-   Fish-friendly: If observed survival ≥ 98% and lower 95% CI ≥ 94%

-   Continue testing: Otherwise

The continue zone therefore directly controls the estimation error and we do not apply traditional hypothesis based testing.\
\
**Accepting only when results are clear and conservative (\>98% + LB ≥ 94%). Rejecting only when results are definitively bad (UB \< 98%). Continuing otherwise—no gray-area decisions.**

**Type I/II Error Control**

-   **Type I Error (False "Accept"):**

    -   Controlled by requiring **LB ≥ 94% AND observed \> 98%**.

    -   This is stricter than classic hypothesis tests (reduces false positives).

-   **Type II Error (False "Reject"):**

    -   Controlled by **only rejecting if UB \< 98%** (avoids premature rejection)

**Conservatism**

-   Sequential testing is required in our application as we do not have sufficient knowledge of the actual mortality rate to apply a fixed sample size. 

-   Our first sample occurs at 100, which has already significantly reduced our likelihood of committing a type I error. We simply maintain our artificial alpha spend linearly throughout the experiment. Bare in mind there is no true alpha when we are just examining within group probability and want a fixed precision. That will come in the next approach of comparing impact to control groups.

-   The criteria for rejecting the null are already high by requiring an upper bounds to not include 98.

-   Fixed lower bound rule is a conservative measure to guarantee fixed estimation lower bounds.

Because this restriction is already conservative, we choose not to apply any further adjustment to the confidence intervals.

## 2.1 Generate mortality data

Construct the mortality table using a max sample size, desired survival and acceptable lower bound estimate. 

95% CI = Wilson binomial. R package 'binom'
Key calculation:

```{r eval=FALSE}
conf <- binom.confint(n_mortality, n, methods = "wilson")

conf <- binom.confint(200, 200, methods = "wilson")
conf
```

```{r mortality_table, eval = FALSE, echo = FALSE}

mortality_data <- create_mortality_table(
  max_sample_size = 1000,
  step_size = 1,
  lower_threshold = 96,
  desired_surv = 98,
  max_mortality_proportion = 0.5
)

#using step size of 1 creates mortality table robust to any recapture rate

write.csv(mortality_data, "./output/mortality_data.csv", row.names = FALSE)

mortality_sample_plot <- decision_plot(mortality_data)
mortality_sample_plot

ggsave(filename = "./output/mortality.svg", plot = mortality_sample_plot,
     units="cm",width=14,height=14)

```

## 2.2 Simulate trial outcomes

Where recapture rate < 100% we need a method to estimate mortality across our observed (recaptured) and unobserved (missing) fish.

Horvitz–Thompson estimator

Key assumption of H-T estimator:

The HT estimator is designed for unbiased estimation under unequal probability sampling, which aligns well with:

 - Known recapture rate (i.e., sampling probability for each fish)

 - A known total number of exposed individuals (the finite population correction, FPC)
 
 - The assumption that mortality is independent of recapture (i.e., missingness is random or at least conditionally ignorable). If unrecaptured fish may have higher or lower mortality, then the HT estimator is biased because the missingness is not at random. This is a core limitation: HT assumes known inclusion probabilities and that these are not related to the variable of interest (here, mortality).

In reality, we may violate this assumption, but we have no prior information to reasonably assume this. Therefore, we can only assume that mortality is at random.

However... H-T provides a defensible first-pass estimate, especially if paired with finite population control (we always have this!)

How may we handle the violation?

Effective sensitivity analysis.

The H-T confidence interval says how uncertain we are in the mortality estimate when we have only observed part of the population, assuming recaptured fish are a random sample. The lower bound of the CI reflects the minimum plausible mortality, the upper bound reflects the maximum plausible mortality.

We accept that the CI bounds assume random sampling, but we apply the lower and upper bounds to allow for uncertainty due to incomplete observation. We do not believe there is a strong bias (assuming random), but want a defensible lower-end estimate.

The CI is derived from normal approximation.

For example:

300 fish exposed
270 recaptured
27 deaths observed, observed mortality = 27/270 = 10%
HT point estimate = 27/270 x 300 = 30 deaths
HT 95% CI: (26.6, 33.4) = 26.6/300 (8.87% mortality), = 33.4/300 (11.13% mortality)

So, under the assumption that the recaptured fish are representative (e.g., assuming random mortality), the true population mortality rate is likely between 8.9% and 11.1% with 95% confidence.

We do not know mortality is random, so we can infer the lower and upper mortality estimates as potentially representing scenarios where mortality is biased in those groups (e.g., lower/higher mortality in missing fish). In this scenario, the lower bounds are always going to represent mortality of 0 in the missing fish.

What do the bounds potentially imply about missing fish?

Implied death in missing fish:

Lower = 26.6 - 27 = -0.4 (not possible death, so assume 0) e.g., 10% percentage points less mortality than recaptured)
Upper = 33.4 - 27 = 6.4 = 6.4/30 = 21.3% (e.g., 11.13% percentage points more morality than recaptured)

These lower and upper bounds are therefore inferential of what COULD'VE happened to the fish. More or less fish may have died in the missing group.

Under the assumptions of the HT estimator, and based on the confidence interval of total mortality, mortality among unrecaptured fish could plausibly range from near 0% (if all survived) to 21.3% (if 6.4 of 30 died). This reflects the uncertainty due to both sampling variation and the unknown fate of missing fish, but does not account for systematic bias if missingness is not random.

Key calculation: R package 'survey'

```{r eval= FALSE}
 # Create data frame for survey design
  recaptured_data <- data.frame(
    death = c(rep(1, observed_deaths), rep(0, observed_survivors)),
    weight = 1 / recapture_rate,
    fpc = exposed_fish
  )
  
  # Survey design with finite population correction
  design <- svydesign(
    ids = ~1,           
    weights = ~weight,  
    fpc = ~fpc,        
    data = recaptured_data
  )
  
  # Calculate HT estimate
  ht_total <- svytotal(~death, design)
  ht_ci <- confint(ht_total)

```

Simulate sequential trial decisions using Horvitz–Thompson estimator.

Three levels of estimation
    aggressive = accept earlier (lowest death estimate mortality lower in missing fish)
    neutral = marginal inflation of lower bounds (mortality assumed equal in missing fish)
    conservative = reject earlier (highest death estimate mortality higher in missing fish)

```{r horvitz–hompson_estimator, eval = FALSE}

execution_time <- system.time(
  ht_simulation_results <- simulate_all_HT(
    simulations = 1:1000,
    max_sample = 1000,
    true_survival_rates = round(seq(0.90, 1, 0.01), 2),
    check_intervals = c(100, 200, 300, 400, 500, 600, 700, 800, 900, 1000),
    mortality_table = mortality_data,
    recapture_rates = round(seq(0.90, 1, 0.01), 2)
  )
)

print(execution_time)

# Takes ~ 2 hours

write.csv(ht_simulation_results, "./output/ht_simulation_results.csv")

```

## 2.3 Import

Import pre-calculated mortality data and simulation results.

Mortality data:
Desired survival 98%
Lower bounds acceptance 96%

Simulation data:
  Max sample size = 1000
  Simulations (for each interval) = 1000
  true_survival_rates = 90 - 100%
  check_intervals = 100 - 1000, by 100
  recapture_rates = 90 - 100%

```{r input_import, echo = FALSE}

mortality_data  <- read.csv("./input/mortality_data_98surv_96lower_step1.csv")
ht_simulation_results  <- read.csv("./input/ht_simulation_results.csv")

#Create demo dataset
#demo_mortality <- mortality_data[sample(nrow(mortality_data), 100), ]
#demo_simulation <- ht_simulation_results[sample(nrow(ht_simulation_results), 100), ]
#write.csv(demo_mortality, "./output/demo_mortality.csv")
#write.csv(demo_simulation, "./output/demo_simulation.csv")

```


## 2.4 Useage

### 2.4.1 Interactive

Create interactive plot for decision making.

```{r plotly_decision_plot, echo=FALSE, eval = FALSE}

decision_point <- mortality_sample_plot +
  geom_point(data = mortality_data, 
             aes(x = sample_size, y = n_mortality, 
                 text = paste("Sample size:", sample_size, "<br>", 
                             survival_formatted),  color = decision),
             alpha = 0, size = 1)+
  scale_color_manual(
    values = c(
      "Accept H0" = "darkgreen", 
      "Reject H0" = "darkred", 
      "Continue" = "grey50"
    )
  )

decision_point$layers <- decision_point$layers[
  !sapply(decision_point$layers, function(x) inherits(x$geom, "GeomText"))
]

decision_point_plotly <- ggplotly(decision_point, tooltip = "text")
tmp <- tempfile(fileext = ".html")
htmlwidgets::saveWidget(decision_point_plotly, tmp)
browseURL(tmp)

```


### 2.4.2 Lookup mortality

```{r lookup_mort, eval = FALSE}
# lookup specific rows (sample, mortality)
lookup_mortality_data(mortality_data, 400, 7) 
lookup_mortality_data(mortality_data, 360, 7) 

# need to add lookup for simulation results too, so we can quickly identify decision points
```

### 2.4.3 Sequential decision

```{r seq_decision, eval = FALSE}

# Plot a specific simulation scenario when working live
# e.g.,  assumed true survival is 95% at interval 300, what do sequential test routes to rejecting null hypothesis this look like? 
# At 300 fish 
# We have a cumulative re-catch of 278 (22 missing fish)
# We have observed 7 dead fish 

# Survival probability of recaptured fish
#271/278 = 97.48%
# Recapture rate
#278/300 = 92.7%
# These show which lookup conditions to input for the simulation

#We then measure mortality with 3 methods and perform sensitivity analysis
#Run with defaults first, then can input the assumed_true_survival (0.98) and recapture_rate_scenario (1) afterwards

my_decisions <- lookup_seq_95(
  exposed_fish = 300,
  recaptured_fish = 270, 
  observed_deaths = 4,
  mortality_table = mortality_data,
  cumulative_data = cumulative_all
)

#needs to 
# show stopping proportions plot
# show simulated trials plot
# show sensitivity analysis plot

```


## 2.5 Visualise trial outcomes

Using the simulation output, we can now determine what scenarios are most likely to occur during sequential testing. This allows us to determine our stopping rate at given sample itnervals and given mortality estimates.

```{r simulation_plot, eval=FALSE}

# Plot simulations for 90 - 100% survival, 90% and 100% recapture using the neutral method. Mortality in missing fish assumed to be random (like observed fish)

journey_plot_recapture <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c(0.90, 0.91, 0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 0.98, 0.99,  1),
    recapture_rates_to_show = c(0.9, 1.0),
  decision_methods_to_show = c("neutral"),
  n_journeys = 1000
)

journey_plot_recapture$layers <- journey_plot_recapture$layers[
  !sapply(journey_plot_recapture$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recapture

```


```{r simulation_plot_examples, eval=FALSE}

journey_plot_recapture <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c(0.90, 0.95),
    recapture_rates_to_show = c(1.0),
  decision_methods_to_show = c("neutral"),
  n_journeys = 1000
)

journey_plot_recapture$layers <- journey_plot_recapture$layers[
  !sapply(journey_plot_recapture$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recapture <- journey_plot_recapture +
  labs(title = NULL, subtitle = NULL)

ggsave(filename="journey_plot_recapture.svg", plot=journey_plot_recapture,
       device = "svg",units="cm", width=24,height=16, dpi=300)

journey_plot_recapture1 <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c(0.97, 0.98),
    recapture_rates_to_show = c(1.0),
  decision_methods_to_show = c("neutral"),
  n_journeys = 1000
)

journey_plot_recapture1$layers <- journey_plot_recapture1$layers[
  !sapply(journey_plot_recapture1$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recapture1 <- journey_plot_recapture1 +
  labs(title = NULL, subtitle = NULL)

ggsave(filename="journey_plot_recapture1.svg", plot=journey_plot_recapture1,
       device = "svg",units="cm", width=24,height=16, dpi=300)


journey_plot_recapture2 <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c(0.99, 1),
    recapture_rates_to_show = c(1.0),
  decision_methods_to_show = c("neutral"),
  n_journeys = 1000
)

journey_plot_recapture2$layers <- journey_plot_recapture2$layers[
  !sapply(journey_plot_recapture2$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recapture2 <- journey_plot_recapture2 +
  labs(title = NULL, subtitle = NULL)

journey_plot_recapture2

ggsave(filename="journey_plot_recapture2.svg", plot=journey_plot_recapture2,
       device = "svg",units="cm", width=24,height=16, dpi=300)

# Plot simulations for specific survival and recapture rate, three mortality estimates

journey_plot_recap_method <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c( 0.98),
    recapture_rates_to_show = c(0.90),
  decision_methods_to_show = c("neutral", "aggressive", "conservative"),
  n_journeys = 1000
)

journey_plot_recap_method$layers <- journey_plot_recap_method$layers[
  !sapply(journey_plot_recap_method$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recap_method

```

Need to add ability to plot specific simulations e.g.,

We're at interval 300
- what do scenarios which end here look like?
- what do scenarios which end at the next interval look like?

## 2.6 Stopping decisions

Now, we need to understand when we are most likely to stop, at a given sample interval, for a given true survival rate and recapture 

```{r create_cumulative}

ht_neutral <- ht_simulation_results %>% 
  filter(decision_method == "neutral")

ht_aggressive<- ht_simulation_results %>% 
  filter(decision_method == "aggressive")

ht_conservative <- ht_simulation_results %>% 
  filter(decision_method == "conservative")

cumulative_data_neutral <- cumulative_stopping(ht_neutral)
cumulative_data_conservative<- cumulative_stopping(ht_conservative)
cumulative_data_aggressive<- cumulative_stopping(ht_aggressive)

# Combine and add method labels to create complete lookup table for stopping decisions
cumulative_all <- bind_rows(
  cumulative_data_aggressive %>% mutate(decision_method = "aggressive"),
  cumulative_data_neutral %>% mutate(decision_method = "neutral"),
  cumulative_data_conservative %>% mutate(decision_method = "conservative")
)

#write.csv(cumulative_all, "./output/cumulative_all.csv")

```

```{r stopping_example, eval=FALSE}

#use lookup table to find specific example of stopping point for a given scenario 
# How many have stopped and accepted, rejected or continued by now using a given method?
# Lookup from cumulative table

stopping_93_recap_all_methods <- cumulative_all %>%
  filter(
    true_survival_rate == 0.98,
    recapture_rate == 0.93
  ) %>%
  select(
    decision_method,
    interval,
    prop_accept_by_now,
    prop_reject_by_now,
    prop_continue,
    prop_stopped_by_now
  ) %>%
  arrange(decision_method, interval)

#Alternatively, we can look just at how many trials have stopped using a specific method
# Given survival, recapture rate, how many trials have stopped by a given sample interval?
stopping_93_recap_method <- cumulative_all %>%
  filter(
    true_survival_rate == 0.98,
    recapture_rate == 1
  ) %>%
  select(interval, decision_method, prop_stopped_by_now)

# Pivot to wide format for easy comparison
stopping_93_recap_method <- stopping_93_recap_method %>%
  pivot_wider(names_from = decision_method, values_from = prop_stopped_by_now)

```


### 2.6.1 Inconclusivity rates

Assuming a max interval size of 1000

```{r inconclusivity_by_method, eval=FALSE}

inconclusivity_by_method <- cumulative_all %>%
  filter(interval == 1000) %>%
  select(true_survival_rate, recapture_rate, decision_method, prop_continue) %>%
  pivot_wider(names_from = recapture_rate, 
              values_from = prop_continue,
              names_prefix = "recapture_") %>%
  mutate(across(starts_with("recapture_"), ~ round(.x * 100, 1))) %>%
  arrange(decision_method, true_survival_rate)

inconclusivity_by_method

```

### 2.6.2 Visualise stopping decisions

We can now visualise our stopping decisions to provide a visual reference for when continuing sampling is a worthy trade-off

```{r plot_cumulative_stopping, eval=FALSE}

# Plot all survival and recapture conditions as a facet matrix 

plot_cumulative_stopping(cumulative_all, 
                         decision_method_filter = "neutral")

#decision_method_filter = "conservative"
#decision_method_filter = "aggressive"

```

```{r eval=FALSE}

#Plot specific scenario showing difference across recapture method
stopping_plot2 <- plot_cumulative_stopping(cumulative_all,
                                          survival_filter = 0.98, 
                                          recapture_filter = 0.9)
stopping_plot2


ggsave(filename="stopping_plot2.svg", plot=stopping_plot2,
       device = "svg",units="cm", width=24,height=10, dpi=300)

#Plot needs to be arranged correctly

```

Make sure all our plot methods arrange the recapture method as aggressive, netural, conservative

## 2.7 Visualise Type I error

We can examine the influence of our recapture estimate on type I error.

```{r cumulative_plot, eval=FALSE}

#Examine acceptance rate across all

cumulative_plot <- ggplot(cumulative_all, 
                         aes(x = interval, y = prop_accept_by_now, 
                             color = decision_method, linetype = decision_method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("aggressive" = "blue", "neutral" = "black", "conservative" = "red")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = c(100, 200, 300, 400, 500, 600, 700, 800, 900, 1000)) +
  labs(
    title = "Cumulative Acceptance Rate by Sample Interval and Decision Method",
    x = "Sample Interval",
    y = "Cumulative Proportion Accepting H0",
    color = "Decision Method",
    linetype = "Decision Method"
  ) +
  theme_JN() +
  facet_grid(recapture_rate ~ true_survival_rate, 
             labeller = labeller(recapture_rate = function(x) paste("Recapture:", x),
                               true_survival_rate = function(x) paste("Survival:", x)))

cumulative_plot


```

```{r cumulative_plot_9398, eval=FALSE}

#Examine acceptance rate across all methods for specific survival and recapture

cumulative_plot_9398 <- cumulative_all %>% 
  filter(true_survival_rate == 0.98, recapture_rate == 0.90) %>%
  select(decision_method, interval, prop_accept_by_now, prop_reject_by_now, prop_continue) %>%
  pivot_longer(cols = c(prop_accept_by_now, prop_reject_by_now, prop_continue),
               names_to = "outcome", 
               values_to = "proportion") %>%
  mutate(
    outcome = case_when(
      outcome == "prop_accept_by_now" ~ "Accept H0",
      outcome == "prop_reject_by_now" ~ "Reject H0", 
      outcome == "prop_continue" ~ "Continue"
    )
  ) %>%
  ggplot(aes(x = interval, y = proportion, 
             color = decision_method, linetype = decision_method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("aggressive" = "blue", "neutral" = "black", "conservative" = "red")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = c(100, 200, 300, 400, 500, 600, 700, 800, 900, 1000)) +
  labs(
    title = "Cumulative Trial Outcomes by Sample Interval (98% Survival, 93% Recapture)",
    x = "Sample Interval",
    y = "Cumulative Proportion",
    color = "Decision Method",
    linetype = "Decision Method"
  ) +
  theme_JN() +
  theme(legend.position = "bottom") +
  facet_wrap(~outcome, scales = "free_y") +
  coord_cartesian(clip = "off")

cumulative_plot_9398


ggsave(filename="cumulative_plot_9398.svg", plot=cumulative_plot_9398,
       device = "svg",units="cm", width=24,height=10, dpi=300)


```


We may then consider two things (1) Where the 'grey zone' exists when we reject null hypothesis and it may actually be true (e.g., the level of acceptance in 0.94 - 0.97). This is our type I error rate, which increase when trying to estimate close to thresholds.
(2) how we can control type I error if we use neutral method as it influences acceptance the least.

```{r sensitivity_plot, eval=FALSE}

ht_summary <- create_stopping_summary_ht(ht_simulation_results)

# Show how decision method affects outcomes across recapture rates
sensitivity_data <- ht_summary$trial_overview %>%
  select(true_survival_rate, recapture_rate, decision_method, prop_accept, prop_continue)

sens_plot <- ggplot(sensitivity_data, aes(x = recapture_rate, y = prop_accept, 
                            color = decision_method, linetype = decision_method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("aggressive" = "blue", "neutral" = "black", "conservative" = "red")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Acceptance Rate by Recapture Rate and Decision Method (98% survival)",
    x = "Recapture Rate",
    y = "Proportion Accepting H0",
    color = "Decision Method",
    linetype = "Decision Method"
  ) +
  theme_JN() +
  facet_wrap(~true_survival_rate)

sens_plot

```


```{r eval=FALSE}

# For a specific example e.g., 98% true survival

sens_plot2 <- ggplot(sensitivity_data %>%
                       filter(true_survival_rate == 0.98), aes(x = recapture_rate, y = prop_accept, 
                            color = decision_method, linetype = decision_method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("aggressive" = "blue", "neutral" = "black", "conservative" = "red")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Acceptance Rate by Recapture Rate and Decision Method (98% survival)",
    x = "Recapture Rate",
    y = "Proportion Accepting",
    color = "Decision Method",
    linetype = "Decision Method"
  ) +
  theme_JN() +
  coord_cartesian(clip = "off")

sens_plot2

ggsave(filename="sens_plot2.svg", plot=sens_plot2,
       device = "svg",units="cm", width=16,height=10, dpi=300)


# We can see that the effect is worst when recapture rate is lower - it reflects our real world confidence!

```


#Need function for finding minimum sample size for given proportion and confidence interval estimate 

# 3. Hypothese testing

## 3.1 Compliance testing

Scenario: Observed 100% survival
          Observed 99% survival
          Observed 97% survival
          Observed 96% survival
          
Objective: Does pump meet regulatory threshold (≥98%)?

An Exact Binomial test can be applied to determine if an observed survival rate is significantly higher (greater) or lower (less) than a given compliance threshold (p = 0.98).The test can be applied where observed survival is 99% or 100% and statistical evidence to support the survival estimate is desired without a control group. When survival is marginally compliant, or non-compliant, the test may be applied but statistical comparison to a control group is preferred in these scenarios.

Excessively large sample sizes will be required to show 99% is > 98%, and 97% < 98%. This is unethical, and should therefore use confidence interval estimation and/or inferiority (impact survival < control survival) or superiority (impact mortality > control mortality) testing.

### 3.1.1 Test

```{r compliance_test, eval= FALSE}

#Superiority approach

# Example data: 100% survival at minimum sample size of 188 where 95% CI = 98.4% - 100%.
binom.test(x = 188, n = 188, p = 0.98, alternative = "greater")
#p = 0.02241

# Example data: 99% survival at minimum sample size of 200 where 95% CI = 96.43% - 99.73%.
binom.test(x = 198, n = 200, p = 0.98, alternative = "greater")
#p = 0.2351
#See power analysis below for required sample size to show 99% exceeds 98% compliance

#Inferiority approach

# Example data: 96% survival at minimum sample size of 200 where 95% CI = 92.31% - 97.96%
binom.test(x = 192, n = 200, p = 0.98, alternative = "less")
#p = 0.04934

# Example data: 97% survival at minimum sample size of 600 where 95% CI = 95.31 - 98.09%
binom.test(x = 582, n = 600, p = 0.98, alternative = "less")
#p = 0.06104
#See power analysis below for required sample size to show 97% does not meet 98% compliance

# check what this power measurement is doing and decide whether to include it
# Power when true survival = 99%
# power <- 1 - pbinom(957, 969, 0.99) 
# power

```

### 3.1.2 Power

```{r compliance_power, eval = FALSE}

result_greater <- find_compliance_sample_size(
  true_rate = 0.99, 
  test_direction = "greater", 
  target_power = 0.8
)
print(result_greater)
# Sample size of at least 969 required to observe an exact 99% survival that statistically exceeds 98%
# Reject (compliance ≤  0.98 ) if ≥  957 / 969  survivors
# Example data: 99% survival at minimum sample size of 200 where 95% CI = 96.43% - 99.73%.
# binom.test(x = 957, n = 969, p = 0.98, alternative = "greater")
# p = 0.04973

result_less <- find_compliance_sample_size(
  true_rate = 0.97, 
  test_direction = "less", 
  target_power = 0.8
)
print(result_less)
# Sample size of at least 1469 required to observe an exact 97% survival that statistically does not meet 98%
# Reject (compliance ≥  0.98 ) if ≤  1430 / 1469  survivors
# Example data: 97% survival at minimum sample size of 600 where 95% CI = 95.31 - 98.09%
# binom.test(x = 1430, n = 1469, p = 0.98, alternative = "less")
# p = 0.04931

```


What about when we observe 98% survival?

Exact binomial test for compliance cannot be applied when observed survival = 98%. Instead, we rely on CI estimation, and can also make interpretation from comparisons with a control group.

Group comparison test directional considerations
  - Two-sided: Is there any statistical difference between pump and control?
  - One-sided (less): Is pump survival inferior to control survival?
      matrix(c(impact_surv, impact_died, control_surv, control_died), nrow=2, byrow=T)
      fisher.test(matrix, alternative = "less")
  - One-sided (greater): Is control survival superior to impact survival?
      matrix(c(control_surv, control_died, impact_surv, impact_died), nrow=2, byrow=T)
      fisher.test(matrix, alternative = "greater")
  - Use a survival-first matrix. It is critical the first row of the matrix represents the direction of compliance
  e.g., impact first = alternative tests against control
  
For consistency, there is no need to consider reversing the matrix and asking if control is superior. We should always determine whether pump survival is inferior. If the two observed proportions are close (e.g., 94% impact, 95% control), we can switch to a TOST but this will hugely increase sample demands.
  
To examine the question of 'Is pump acceptable despite meeting threshold exactly?', we can take a difference or inferiority test approach. The same approach will also be followed for determining test choice where we ask 'Does the pump cause harm?'

## 3.2 Difference testing

### 3.2.1 Test

- Two-sided: Is there any statistical difference between pump and control?

Scenario: Observed 98% survival
          Observed ≤ 97% survival

Objective: Is there any difference between pump and control?

A two-sided Fisher's Exact test can be applied to determine if an observed survival rate is significantly different (non-directional) than a control survival. The test can be applied where observed survival is ≤ 98% and statistical evidence to support the survival estimate is desired with a control group.

Excessively large sample sizes will be required to show 98% is different to 100%. This is unethical, and should therefore use confidence interval estimation primarily as little is gained by showing difference with a control group.

```{r difference_test, eval=FALSE}

# Example data 98% survival at minimum sample size of 400 where 95% CI = 96.11% - 98.98%
observed_impact_surv = 0.98
impact_n = 400
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 134 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.219
#See power analysis below for required sample size to show a difference between impact (98%) and control (100%) exists


# Example data: 97% survival at minimum sample size of 600 where 95% CI = 95.31 - 98.09%
observed_impact_surv = 0.97
impact_n = 600
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 200 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.01022
# Suggests sample size could actually be lower than if using CI alone
#See power analysis below for required sample size to show a difference between impact (97%) and control (100%) exists


# Example data: 90% survival at minimum sample size of 100 where 95% CI = 82.56 -  94.48%
observed_impact_surv = 0.97
impact_n = 100
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 33 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.5738
#See power analysis below for required sample size to show a difference between impact (90%) and control (100%) exists


```

### 3.2.1 Power

```{r difference_power, eval = FALSE}

# Test for any difference between groups when impact survival = 98%
find_min_sample_size_fisher(p_impact = 0.98, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "two.sided")

# sample size of at least impact n = 735, control n = 245 to  show a difference between impact (98%) and control (100%) exists

# observed_impact_surv = 0.98
# impact_n = 735
# surv_impact <- observed_impact_surv*impact_n
# dead_impact <- impact_n - surv_impact

# observed_control_surv = 1
# control_n = 245 #approx 1:3
# surv_control <- observed_control_surv*control_n
# dead_control <- control_n - surv_control

# survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
#                              nrow = 2, byrow = TRUE)
# fisher.test(survival_matrix, alternative = "two.sided")
#p = 0.02944

# Test for any difference between groups when impact survival = 97%
find_min_sample_size_fisher(p_impact = 0.97, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "two.sided")

# sample size of at least impact n = 492, control n = 164 to  show a difference between impact (97%) and control (100%) exists

 # observed_impact_surv = 0.97
 # impact_n = 492
 # surv_impact <- observed_impact_surv*impact_n
 # dead_impact <- impact_n - surv_impact
 # 
 # observed_control_surv = 1
 # control_n = 164 #approx 1:3
 # surv_control <- observed_control_surv*control_n
 # dead_control <- control_n - surv_control
 # 
 # survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
 #                              nrow = 2, byrow = TRUE)
 # fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.02885

# Test for any difference between groups when impact survival = 90%
find_min_sample_size_fisher(p_impact = 0.9, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "two.sided")

# sample size of at least impact n = 147, control n = 49 to  show a difference between impact (90%) and control (100%) exists

 # observed_impact_surv = 0.9
 # impact_n = 147
 # surv_impact <- observed_impact_surv*impact_n
 # dead_impact <- impact_n - surv_impact
 # 
 # observed_control_surv = 1
 # control_n = 49 #approx 1:3
 # surv_control <- observed_control_surv*control_n
 # dead_control <- control_n - surv_control
 # 
 # survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
 #                              nrow = 2, byrow = TRUE)
 # fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.0247


```

Powering the test to detect a difference is not statistically meaningful when we observe marginally compliant survival, or where control survival is imperfect and we need to understand if the pump was actually harmful. 

## 3.3 Inferiority testing

### 3.3.1 Test

- One-sided: Is impact survival statistically lower than control survival?

Scenario: Observed 98% survival
          Observed ≤ 97% survival

Objective: Is pump survival inferior to control survival? Does the pump cause harm?

A one-sided Fisher's Exact test can be applied to determine if an observed impact survival rate is significantly lower (less) than a control survival. The test can be applied where observed survival is ≤ 98% and statistical evidence to support the survival estimate is desired with a control group.

What about when we observe non-compliant pump survival but control survival was also < 100%?

The objective changes but the test remains the same.

Objective: Does the pump harm fish more than background mortality (e.g., mortality caused by net or handling control)?
           Prove the pump is worse for fish that enter the pump than fish which do not enter the pump. Pump causes demonstrable harm.

The test choice does not change as we are examining absolute % difference between impact and control. What is important is the % difference between the two samples, not the overall proportion e.g., 47% impact vs 50% control is the same as 97% impact vs 100% control. However, it can theoretically detect any difference, and a significant result could therefore falsely conclude pump is harmful if there is enough statistical power to detect a small difference (e.g., 1%). Therefore, this means interpretation of the test results must accept that non-significant ≠ pump is safe but instead means insufficient statistical evidence to show harm. 

•	Null hypothesis (H0) Survival in the pump exposure is not less than survival in the control exposure
•	Alternative hypothesis (H1) Survival in the pump exposure is less than survival in the control exposure

```{r inferiority_test, eval = FALSE}

# Example data 98% survival at minimum sample size of 400 where 95% CI = 96.11% - 98.98%
observed_impact_surv = 0.94
impact_n = 243
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 121 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "less")

# p = 0.09737
#See power analysis below for required sample size to show impact (98%) survival is less than control (100%).


# Example data: 97% survival at minimum sample size of 600 where 95% CI = 95.31 - 98.09%
observed_impact_surv = 0.97
impact_n = 600
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 200 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "less")

#p = 0.005284
# Suggests sample size could actually be lower than if using CI alone
#See power analysis below for required sample size to show a difference between impact (97%) and control (100%) exists


# Example data: 90% survival at minimum sample size of 100 where 95% CI = 82.56 -  94.48%
observed_impact_surv = 0.97
impact_n = 100
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 33 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "less")

#p = 0.4219
#See power analysis below for required sample size to show a difference between impact (90%) and control (100%) exists


```
The key observation is that we increase statistical power when using a one-sided test. We are able to justify this as the pump is never likely to exceed control survival, nor is significance in this direction of interest.

### 3.2.1 Power

Power analysis is used to simulate the observed data with a nominal number of simulations (e.g., 1000) to determine the power to detect a given alpha (e.g., 0.05). This means, every result in the simulation can theoretically achieve an alpha < 0.05, but the power reflects how likely we are to detect a significant result. Therefore, low powered studies are unreliable as we cannot confidently conclude that we would observe the same effect again, if we gathered the data from another independent population.

Importance of correct study power
1. We find a p-value = 0.01. This means, if there was truly NO difference, there's only a 1% chance I'd see results this extreme by random luck.
2. Power of the study was 46%. This means, if there IS a real difference, I'd only detect it 46% of the time.

Therefore, low powered studies are likely to increase Type I error rates e.g., Reject the null hypothesis when it is actually true. In our inferiority scenario, this means we can falsely conclude that pump exposure is less than survival, when actually it is not and a higher powered study would not have found this effect. We conclude the pump is harmful when actually the study was under powered and we made a false conclusion.

Therefore we need two considerations:
1. What does the power curve look like for each survival difference? This shows all the possible significant results and the respective power to detect them at a given sample size.
2. What does the 80% power curve look like across each survival difference? This shows the minimum sample size required to achieve 80% power

First, we can simulate specific scenarios for indivdual assesments.

```{r inferiority_power, eval = FALSE}

# Test for any difference between groups when impact survival = 98%
find_min_sample_size_fisher(p_impact = 0.99, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "less")

# sample size of at least impact n = 681, control n = 227 to show impact (98%) is inferior to control (100%)

 # observed_impact_surv = 0.98
 # impact_n = 681
 # surv_impact <- observed_impact_surv*impact_n
 # dead_impact <- impact_n - surv_impact
 # 
 # observed_control_surv = 1
 # control_n = 227 #approx 1:3
 # surv_control <- observed_control_surv*control_n
 # dead_control <- control_n - surv_control
 # 
 # survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
 #                              nrow = 2, byrow = TRUE)
 # fisher.test(survival_matrix, alternative = "less")
#p = 0.01723

# Test for any difference between groups when impact survival = 97%
find_min_sample_size_fisher(p_impact = 0.97, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "less")

# sample size of at least impact n = 453, control n = 151 to show impact (97%) is inferior to control (100%) exists

 # observed_impact_surv = 0.97
 # impact_n = 453
 # surv_impact <- observed_impact_surv*impact_n
 # dead_impact <- impact_n - surv_impact
 # 
 # observed_control_surv = 1
 # control_n = 151 #approx 1:3
 # surv_control <- observed_control_surv*control_n
 # dead_control <- control_n - surv_control
 # 
 # survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
 #                              nrow = 2, byrow = TRUE)
 # fisher.test(survival_matrix, alternative = "less")

#p = 0.01693

# Test for any difference between groups when impact survival = 90%
find_min_sample_size_fisher(p_impact = 0.9, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "less")

# sample size of at least impact n = 135, control n = 45 to show impact (90%) is inferior to control (100%) exists

 observed_impact_surv = 0.9
 impact_n = 135
 surv_impact <- observed_impact_surv*impact_n
 dead_impact <- impact_n - surv_impact

 observed_control_surv = 1
 control_n = 45 #approx 1:3
 surv_control <- observed_control_surv*control_n
 dead_control <- control_n - surv_control

 survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
                              nrow = 2, byrow = TRUE)
 fisher.test(survival_matrix, alternative = "less")

#p = 0.01532


```

Second, we can simulate all difference of 1 - 50% to generate multiple power curves and comprehensive sample planning.

### 3.2.2 Power simulation

```{r inferiority_power_all_diff, eval= FALSE}

# Power Analysis Based on Percentage Point Differences
# Same structure as original but using % differences instead of fixed survivals

# Set parameters
control_sizes <- seq(1, 500, by = 1)
percentage_differences <- seq(1, 50, by = 1)  # 1% to 50% difference
baseline_control_survival <- 1.0  # Control always 100%
alpha <- 0.05
group_ratio <- 3  # 3:1 impact:control

power_results <- expand_grid(
  n_control = control_sizes,
  pct_difference = percentage_differences
) %>%
  mutate(
    n_impact = n_control * group_ratio,
    n_total = n_control + n_impact,
    control_survival = baseline_control_survival,
    impact_survival = baseline_control_survival - (pct_difference / 100),
    # Ensure impact survival doesn't go below 0
    impact_survival = pmax(impact_survival, 0)
  ) %>%
  filter(impact_survival >= 0)  # Remove impossible combinations

# Simulate power using simulate_fisher_power function (number of iterations set in function, currently 1000)
# wrap in execution time to measure processing time

execution_time <- system.time(
power_results$power <- pmap_dbl(
  power_results %>% select(n_impact, n_control, impact_survival, control_survival),
  function(n_impact, n_control, impact_survival, control_survival) {
    simulate_fisher_power(
      n_impact = n_impact,
      n_control = n_control,
      p_impact = impact_survival,
      p_control = control_survival,
      alpha = alpha,
      alternative = "less"
    )
  }
)
)
# simulation runs 1000 tests for each combination of control/impact allocation and percentage difference
# 25,000 combinations = 25 million tests!
# Takes approx 2.5 hours to simulate 25 million tests
#write_csv(power_results, "./input/inferiority_power_results.csv")

```

Import previous simulation data

```{r inferiority_power_import, echo = FALSE}

inferiority_power_results  <- read.csv("./input/inferiority_power_results.csv")

#control survival set as 100%, impact survival 99% - 50%. Labels irrelevant as it represents any % difference between impact and control

```

### 3.2.3 Visualise power curves

```{r inferiority_power_vis}

#Power Curves: Fisher's Exact Test (alternative = 'less'). Testing if impact survival < control survival, 3:1 allocation

# Show power curves for 1% through 10% differences
inferiority_power_plot <- inferiority_power_results %>%
  filter(pct_difference %in% c(1, 2, 3, 6, 8, 10)) %>%
  ggplot(aes(x = n_total, y = power, color = factor(pct_difference))) +
  geom_smooth(se = FALSE, size = 1, method = "loess", span = 0.1) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "black", size = 0.8) +
  scale_x_continuous(
    breaks = seq(0, 2000, by = 100),
    limits = c(100, 2000),
    expand = c(0,0)) +
   scale_y_continuous(
    breaks = seq(0, 1, by = 0.1), 
    limits = c(0, 1), 
    expand = c(0, 0.01),
    labels = scales::percent_format(accuracy = 1) ) +
   scale_color_manual(
    name = "Difference\n(% points)",
    values = c(
      "1" = "#999999",  # Light gray
      "2" = "#D55E00",  # Vermillion
      "3" = "#56B4E9",  # Sky blue
      "6" = "#009E73",  # Bluish green
      "8" = "#0072B2",  # Blue
      "10" = "#CC79A7"  # Reddish purple
    )
  )+
  labs(
    x = "Total Sample Size (Impact:Control 3:1 allocation)",
    y = "Statistical Power"
  ) +
  annotate(
    "text",
    x = 1800, 
    y = 0.9, 
    label = "Fisher's exact test (one-sided)\nPower (1-β) = 80%\nα = 0.05\nSamples = 1000",
    hjust = 1,
    vjust = 1,
    size = 4,
    color = "black",
    fontface = "italic"
  ) +
  theme_JN() +
  theme(
    legend.position = "right",
     panel.grid.major = element_line(color = alpha("black", 0.1), size = 0.5),
      panel.ontop = FALSE,
     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)
    )+
  coord_cartesian(clip = "off")

inferiority_power_plot

```

Interactive plot to show sample sizes across power curves

```{r plotly_decision_plot, echo=FALSE, eval = FALSE}

inferiority_point <- inferiority_power_plot +
  geom_point(data = inferiority_power_results, 
             aes(x = n_total, y = power, 
                 text = paste(
                   "Total size:", n_total, "<br>",
                   "Control size:", n_control, "<br>",
                   "Impact size:", n_impact)
                 ),
             alpha = 0, size = 1)

inferiority_point_plotly <- ggplotly(inferiority_point, tooltip = "text")
tmp <- tempfile(fileext = ".html")
htmlwidgets::saveWidget(inferiority_point_plotly, tmp)
browseURL(tmp)

```

### 3.2.4 Summarise minimal samples

```{r min_inferiority_results}

#Follows 3:1 allocation as previous

min_inferiority_results <- inferiority_power_results %>%
  filter(power >= 0.8) %>%
  group_by(pct_difference) %>%
  slice_min(n_total, n = 1) %>%  # Get minimum sample size achieving 80% power
  ungroup() %>%
  select(pct_difference, n_control, n_impact, n_total, power, 
         control_survival, impact_survival) %>%
  rename(achieved_power = power)

min_inferiority_results

```


### 3.2.5 Visualise minimal samples


```{r min_inferiority_plot}

#Minimum Sample Size for 80% Power & α ≤ 0.05/Earliest N where Fisher's exact test (alternative='less') achieves adequate power

annotation_points <- min_inferiority_results %>%
  filter(pct_difference %in% c(1, 2, 3)) %>%
  mutate(
    label = paste(
      "Total size:", n_total,"\n",
      "Control size:", n_control,
      "Impact size:", n_impact
    )
  )

min_inferiority_plot <- min_inferiority_results %>%
  ggplot(aes(x = pct_difference, y = n_total)) +
  geom_smooth(se = FALSE, size = 1, method = "loess", span = 0.2, color = "black") +
  geom_point(data = annotation_points, aes(x = pct_difference, y = n_total), 
             color = "black", size = 3, shape = 16) +
  geom_label(
    data = annotation_points,
    aes(x = pct_difference, y = n_total, label = label),
    color = "black",
    size = 3,
    hjust = -0.1,
    vjust = 0.5,
    fill = NA,
    label.size = 0
  ) +
    annotate(
    "text",
    x = 18, 
    y = 1800, 
    label = "Fisher's exact test (one-sided)\nPower (1-β) = 80%\nα = 0.05\nSamples = 1000",
    hjust = 1,
    vjust = 1,
    size = 4,
    color = "black",
    fontface = "italic"
  ) +
  scale_x_continuous(breaks = seq(0, 20, by = 1), limits = c(0,20), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(0, 1800, by = 100), limits = c(0, 1900), expand = c(0,0)) +
  labs(
    x = "Difference control - impact (% points)",
    y = "Total Sample Size (Impact:Control 3:1 allocation)",
  ) +
  theme_JN() +
  theme(
    panel.grid.major = element_line(color = alpha("black", 0.1), size = 0.5),
    panel.ontop = FALSE
  ) +
  coord_cartesian(clip = "off")

min_inferiority_plot

```

### 3.2.6 Group sequential design

Now we know the fixed sample design requirement to statistically show impact survival is 2% less than control we can construct a group sequential design that allows for early stopping if we observe an increased difference earlier than our total planned sample size.

The purpose of this design is to control our alpha spend throughout planned interim looks at the data. This prevents stopping early without high supportive evidence to make the decision (as it's unknown if stopping earlier prevents observing the true survival).

Two considerations for alpha spending are:
1. O'Brien-Fleming (OBF) Spending
The OBF spending function is highly conservative in the early stages of a trial. It requires an extremely large treatment effect (a very high Z-score) to cross the efficacy boundary at an interim analysis. As the trial progresses, the required Z-score decreases. By the final analysis, the critical value is very close to what it would have been in a fixed-sample design (e.g., 1.96 for a two-sided alpha=0.05). It preserves the overall power of the trial extremely well. The "price" paid for the multiple looks is very small. It's very robust to fluke positive results that might occur by chance at an early interim look.

2. Pocock Spending
The Pocock spending function spends alpha more evenly across all analyses. The Z-score threshold for efficacy is roughly the same at each interim look and the final look (though it's always slightly more strict than 1.96). It is optimized for stopping the trial as soon as possible if the treatment effect is real and substantial. This is its primary benefit. This approach incurs a larger penalty for multiple looks. To maintain the overall Type I error rate, the final analysis requires a p-value stricter than the traditional 0.05 (e.g., a Z-score of ~2.2 instead of 1.96). Pocock is preferred in specific scenarios where the ethical imperatives, financial costs and practical requirements to stop early outweighs the statistical conservatism.

Ultimately, we must make a decision up-front about what the expected survival difference is going to be, as introducing sequential design choice will increase sample size unless we get very strong evidence early that survival difference is greater than expected.

For example, if we believe survival will be 6% difference, we can design a study around a max sample size of ~308, expose all of our fish and collect statistically meaningful data. But, if we planned the study for a 2% difference, we would have to continue until we exceed the sequential stopping boundary.

Required sample size to observe a 6% difference
impact: 231, control: 77 = 308 total
p-value = 0.01609

However, if we planned for 2% difference, and we observed a 6% difference at interval 3 (364 fish stopping point)
Impact: 243, control: 121
p-value = 0.002009
Convert p-value to z-score = Z-score: 2.877 

We can stop early in the trial. But we have used 56 more fish than if we had a fixed sample design. 

This is the real cost of alpha spending to ensure we do not commit Type I error. We require higher sample sizes if a study is designed for a smaller effect size. We simply require stronger evidence to stop earlier in a trial planned for a smaller effect size. This is the intention of group sequential design, we have a penalty for multiple hypothesis testing to control the family wise error rate!

Bottom line = if study designed for smaller effect, we can stop early, but requires more fish than if study designed for larger effect size

Requirements:

One sided Fisher's exact test (H1: p_impact < p_control)
Alpha = 0.05
Beta = 0.20
Spending function O'Brien-Fleming 
Allocation 3:1 impact:control
Interim analyses every 100 fish
Number for a fixed sample design = 908

What is the endpoint design (normal or binomial)?


```{r group_seq_design}

# One-sided, 10 interim analyses, Pocock boundaries based on observing a 2% impact:control difference.
pcock_2perc_design <- gsDesign(
  k = 10,                # Number of interim analyses
  test.type = 1,        # 1 = one-sided (H1: p_impact < p_control)
  alpha = 0.05,         # Type I error
  beta = 0.20,          # Type II error (Power = 80%)
  sfu = "Pocock",            # O'Brien-Fleming spending function
  n.fix = 908 # Fixed sample design informed by power analysis = total sample of 908
)

# One sided test will only present upper boundary for type I error. 
# We do not need lower boundary for futility if we are only interested in detecting a one.sided effect (pump survival less than control) - is this correct?

# Print stopping boundaries (upper bounds are for efficacy)
#print(pcock_2perc_design)

pocock_boundaries_table <- data.frame(
  analysis = 1:10,
  sample_size = ceiling(pcock_2perc_design$n.I),
  z_boundary = round(pcock_2perc_design$upper$bound, 3),
  p_boundary = round(pnorm(pcock_2perc_design$upper$bound, lower.tail = FALSE), 4)
)

pocock_boundaries_table

plot(pcock_2perc_design, plottype = 1)
# Shows our critical threshold of 2.27 and stopping points. Note that if we go to our final 'look', we have increased sample size from 908 to 1214 due to the penalty/alpha spend

#plot(design, plottype = 2)
#plot(design, plottype = 3)
#plot(design, plottype = 4)
#plot(design, plottype = 6)

```

Visualize the stopping boundary/sample size needed
These plots need work, just needs to replicate the plot from gsDesign but with a cleaner appearance

```{r pocock_boundaries_plot}

p_boundaries <- boundaries_table %>%
  ggplot(aes(x = analysis, y = z_boundary)) +
  geom_line(size = 1.5, color = "red") +
  geom_point(size = 4, color = "red") +
  geom_text(aes(label = paste("Z =", z_boundary)), 
            vjust = -0.5, size = 3.5, fontface = "bold") +
  geom_hline(yintercept = 1.96, linetype = "dashed", alpha = 0.6) +
  scale_x_continuous(breaks = 1:10, labels = paste("Analysis", 1:10)) +
  #scale_y_continuous(breaks = seq(1.5, 4, by = 0.5), limits = c(1.5, 4.2)) +
  labs(
    title = "Sequential Stopping Boundaries for 2% Difference",
    subtitle = "Stop if your Z-score exceeds the red line at any analysis",
    x = "Interim Analysis",
    y = "Z-score Boundary (Stop if Z > this value)",
    caption = "Dashed line = traditional fixed study boundary (Z = 1.96)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    panel.grid.minor = element_blank()
  ) +
  annotate("text", x = 3.5, y = 2.1, 
           label = "Fixed study\nboundary", 
           size = 3, alpha = 0.7)

p_boundaries

p_sample_sizes <- boundaries_table %>%
  ggplot(aes(x = analysis, y = sample_size)) +
  geom_line(size = 1.5, color = "blue") +
  geom_point(size = 4, color = "blue") +
  geom_text(aes(label = paste(sample_size, "fish")), 
            vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_x_continuous(breaks = 1:10, labels = paste("Analysis", 1:10)) +
  scale_y_continuous(breaks = seq(0, 1000, by = 100)) +
  labs(
    title = "Sample Sizes for Each Interim Analysis",
    subtitle = "When to conduct each interim look",
    x = "Interim Analysis",
    y = "Total Sample Size"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    panel.grid.minor = element_blank()
  )

p_sample_sizes

```

At each planned stopping interval (given in pocock_boundaries_table), calculate the Fisher's Exact test and check the p value against the p_boundary/z_boundary. This is pre-calculated, but can be checked for specific analysis on the go using p to z calculator

```{r p_to_z}

# Visualise what the stopping p value looks like for the givn z threshold
p_to_z_visualizer(target_z = 2.27)

# Using an example where we are interval 3 (total sample size = 364)
# We have observed 6% survival difference in our impact group, can we stop?

observed_impact_surv = 0.94
impact_n = 243
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 121 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "less")

#p-value = 0.002009

#This exceeds the fixed p value threshold of 0.0116, so we know we can stop
# lets check the Z score anyways

p_to_z_calculator(0.002009)

#Z-score: 2.877 
# We can safely stop the test early and accept our result

```


## 3.5 Equivelance testing

### 3.5.1 Test

Equivalence test: Is pump performance within 2% of background mortality (our new accepted standard based on control background mortality)? This approaches uses TOST but is unlikely to be suitable here because it is a very conservative approach. The test requires the entire confidence interval to fit within the given equivalence bounds. Therefore, sample sizes are likely to be very large and an even higher demand than difference or inferiority testing.

```{r tost_test, eval= FALSE}

# Example data: 90% pump survival and 91% control survival
# Test for equivalence
TOSTtwo.prop(prop1 = 0.9,      # e.g., 0.94
             prop2 = 0.91,   # e.g., 0.96  
             n1 = 800,                 # e.g., 300
             n2 = 800,              # e.g., 200
             low_eqbound = -0.02,             # -2% margin
             high_eqbound = 0.02,             # +2% margin
             alpha = 0.05)

```

### 3.5.1 Power

```{r tost_power, eval = FALSE}

#Not working

# Power analysis
powerTOSTtwo.prop(alpha = 0.05,
                  statistical_power = 0.8,
                  prop1 = 0.9,              # expected impact
                  prop2 = 0.91,              # expected control
                  N = NULL,                  # calculate required N
                  low_eqbound_prop = -0.02,  # -2% margin
                  high_eqbound_prop = 0.02)  # +2% margin

```


Overall, our entire fish survival framework comes down to five considerations.

- 95% Confidence interval estimation
  Always applied to ensure precise survival estimate. Used exclusively for high survival scenarios to prevent unethical sample designs (98%, 99%, 100%)
  
- Exact Binomial compliance test
  May be applied to show compliance if a control group is committed, or when survival is high and desire to     statistically prove > 98%.
  
- Fisher's Exact test (two.sided difference)
  May be applied to show a difference between impact and control fish when survival < 99%. Not advised to apply as we are interested in pump harm.
  
- Fisher's Exact test (one.sided inferiority)
  The primary test consideration when impact survival < 98%. Show inferiority from impact group to control group.
  
- Pocock grouped sequential design
  The primary method for controlling alpha spend with interim analysis.
