---
title: "Performing survival assesments for hydropower and pump compliance: Considerations for test choice, estimation and ethical implications"
author: "Josh Norman"
date: "2025-07-08"
output: html_document
---

# 1. Required libraries

```{r libraries, echo=FALSE}
#library(here)
library(tidyverse)
library(binom)
library(scales)
library(plotly)
library(survey) 
library(htmltools)
library(pwr)
library(TOSTER)
library(cowplot)
library(gsDesign)
#library(gganimate)
#library(ggpubr)
#library(gt)
#library(kableExtra)

#Source functions for calculations, plots etc
source("./functions.R")

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = getwd())
```

# 2. High survival estimation

For high survival scenarios rely on estimation instead of hypothesis-based testing.

This analysis aims to produce a viable Fixed-Width Sequential Confidence Interval Testing method for designing live fish trials with defined points to accept or reject null hypothesis and continue testing.

Fixed-precision Sequential Confidence Interval Testing

Fixed-width sequential confidence interval testing is a statistical method used to estimate a parameter (e.g., mean, proportion) with a predetermined precision while dynamically adjusting the sample size based on accumulating data. Unlike fixed-sample designs, this approach continuously monitors the width of the confidence interval (CI) during data collection, stopping only once the CI narrows to a pre-specified fixed width, ensuring the estimate meets the desired margin of error. Sequential testing is particularly useful when data collection is costly or time-consuming, as it often requires fewer observations than fixed-sample methods while maintaining statistical validity.

We deviate from this principle by fixing our precision level as apposed to the width of the CI.

Ensure we calculate our confidence intervals based on the survival probability, NOT, the mortality probability. This may be a source of error for the previous differences in calculations. 

At each sampling point (e.g., every n = 100), calculate Wilson confidence intervals for observed mortality number. This will inherently deviate from fixed-width CI as the CI width is determined by observed mortality and sample size.

Decision rules to estimate probability with a given precision level (differs from hypothesis based testing):

-   Desired threshold for survival ≥ 98%

-   Not fish-friendly : If upper 95% CI \< 98%

-   Fish-friendly: If observed survival ≥ 98% and lower 95% CI ≥ 94%

-   Continue testing: Otherwise

The continue zone therefore directly controls the estimation error and we do not apply traditional hypothesis based testing.\
\
**Accepting only when results are clear and conservative (\>98% + LB ≥ 94%). Rejecting only when results are definitively bad (UB \< 98%). Continuing otherwise—no gray-area decisions.**

**Type I/II Error Control**

-   **Type I Error (False "Accept"):**

    -   Controlled by requiring **LB ≥ 94% AND observed \> 98%**.

    -   This is stricter than classic hypothesis tests (reduces false positives).

-   **Type II Error (False "Reject"):**

    -   Controlled by **only rejecting if UB \< 98%** (avoids premature rejection)

**Conservatism**

-   Sequential testing is required in our application as we do not have sufficient knowledge of the actual mortality rate to apply a fixed sample size. 

-   Our first sample occurs at 100, which has already significantly reduced our likelihood of committing a type I error. We simply maintain our artificial alpha spend linearly throughout the experiment. Bare in mind there is no true alpha when we are just examining within group probability and want a fixed precision. That will come in the next approach of comparing impact to control groups.

-   The criteria for rejecting the null are already high by requiring an upper bounds to not include 98.

-   Fixed lower bound rule is a conservative measure to guarantee fixed estimation lower bounds.

Because this restriction is already conservative, we choose not to apply any further adjustment to the confidence intervals.

## 2.1 Generate mortality data

Construct the mortality table using a max sample size, desired survival and acceptable lower bound estimate. 

95% CI = Wilson binomial. R package 'binom'
Key calculation:

```{r eval=FALSE}
conf <- binom.confint(n_mortality, n, methods = "wilson")

conf <- binom.confint(200, 200, methods = "wilson")
conf
```

```{r mortality_table, eval = FALSE, echo = FALSE}

mortality_data <- create_mortality_table(
  max_sample_size = 1000,
  step_size = 1,
  lower_threshold = 96,
  desired_surv = 98,
  max_mortality_proportion = 0.5
)

#using step size of 1 creates mortality table robust to any recapture rate

write.csv(mortality_data, "./output/mortality_data.csv", row.names = FALSE)

mortality_sample_plot <- decision_plot(mortality_data)
mortality_sample_plot

ggsave(filename = "./output/mortality.svg", plot = mortality_sample_plot,
     units="cm",width=14,height=14)

```

## 2.2 Simulate trial outcomes

Where recapture rate < 100% we need a method to estimate mortality across our observed (recaptured) and unobserved (missing) fish.

Horvitz–Thompson estimator

Key assumption of H-T estimator:

The HT estimator is designed for unbiased estimation under unequal probability sampling, which aligns well with:

 - Known recapture rate (i.e., sampling probability for each fish)

 - A known total number of exposed individuals (the finite population correction, FPC)
 
 - The assumption that mortality is independent of recapture (i.e., missingness is random or at least conditionally ignorable). If unrecaptured fish may have higher or lower mortality, then the HT estimator is biased because the missingness is not at random. This is a core limitation: HT assumes known inclusion probabilities and that these are not related to the variable of interest (here, mortality).

In reality, we may violate this assumption, but we have no prior information to reasonably assume this. Therefore, we can only assume that mortality is at random.

However... H-T provides a defensible first-pass estimate, especially if paired with finite population control (we always have this!)

How may we handle the violation?

Effective sensitivity analysis.

The H-T confidence interval says how uncertain we are in the mortality estimate when we have only observed part of the population, assuming recaptured fish are a random sample. The lower bound of the CI reflects the minimum plausible mortality, the upper bound reflects the maximum plausible mortality.

We accept that the CI bounds assume random sampling, but we apply the lower and upper bounds to allow for uncertainty due to incomplete observation. We do not believe there is a strong bias (assuming random), but want a defensible lower-end estimate.

The CI is derived from normal approximation.

For example:

300 fish exposed
270 recaptured
27 deaths observed, observed mortality = 27/270 = 10%
HT point estimate = 27/270 x 300 = 30 deaths
HT 95% CI: (26.6, 33.4) = 26.6/300 (8.87% mortality), = 33.4/300 (11.13% mortality)

So, under the assumption that the recaptured fish are representative (e.g., assuming random mortality), the true population mortality rate is likely between 8.9% and 11.1% with 95% confidence.

We do not know mortality is random, so we can infer the lower and upper mortality estimates as potentially representing scenarios where mortality is biased in those groups (e.g., lower/higher mortality in missing fish). In this scenario, the lower bounds are always going to represent mortality of 0 in the missing fish.

What do the bounds potentially imply about missing fish?

Implied death in missing fish:

Lower = 26.6 - 27 = -0.4 (not possible death, so assume 0) e.g., 10% percentage points less mortality than recaptured)
Upper = 33.4 - 27 = 6.4 = 6.4/30 = 21.3% (e.g., 11.13% percentage points more morality than recaptured)

These lower and upper bounds are therefore inferential of what COULD'VE happened to the fish. More or less fish may have died in the missing group.

Under the assumptions of the HT estimator, and based on the confidence interval of total mortality, mortality among unrecaptured fish could plausibly range from near 0% (if all survived) to 21.3% (if 6.4 of 30 died). This reflects the uncertainty due to both sampling variation and the unknown fate of missing fish, but does not account for systematic bias if missingness is not random.

Key calculation: R package 'survey'

```{r eval= FALSE}
 # Create data frame for survey design
  recaptured_data <- data.frame(
    death = c(rep(1, observed_deaths), rep(0, observed_survivors)),
    weight = 1 / recapture_rate,
    fpc = exposed_fish
  )
  
  # Survey design with finite population correction
  design <- svydesign(
    ids = ~1,           
    weights = ~weight,  
    fpc = ~fpc,        
    data = recaptured_data
  )
  
  # Calculate HT estimate
  ht_total <- svytotal(~death, design)
  ht_ci <- confint(ht_total)

```

Simulate sequential trial decisions using Horvitz–Thompson estimator.

Three levels of estimation
    aggressive = accept earlier (lowest death estimate mortality lower in missing fish)
    neutral = marginal inflation of lower bounds (mortality assumed equal in missing fish)
    conservative = reject earlier (highest death estimate mortality higher in missing fish)

```{r horvitz–hompson_estimator, eval = FALSE}

execution_time <- system.time(
  ht_simulation_results <- simulate_all_HT(
    simulations = 1:1000,
    max_sample = 1000,
    true_survival_rates = round(seq(0.90, 1, 0.01), 2),
    check_intervals = c(100, 200, 300, 400, 500, 600, 700, 800, 900, 1000),
    mortality_table = mortality_data,
    recapture_rates = round(seq(0.90, 1, 0.01), 2)
  )
)

print(execution_time)

# Takes ~ 2 hours

write.csv(ht_simulation_results, "./output/ht_simulation_results.csv")

```

## 2.3 Import

Import pre-calculated mortality data and simulation results.

Mortality data:
Desired survival 98%
Lower bounds acceptance 96%

Simulation data:
  Max sample size = 1000
  Simulations (for each interval) = 1000
  true_survival_rates = 90 - 100%
  check_intervals = 100 - 1000, by 100
  recapture_rates = 90 - 100%

```{r input_import, echo = FALSE}

mortality_data  <- read.csv("./input/mortality_data_98surv_96lower_step1.csv")
ht_simulation_results  <- read.csv("./input/ht_simulation_results.csv")

#Create demo dataset
#demo_mortality <- mortality_data[sample(nrow(mortality_data), 100), ]
#demo_simulation <- ht_simulation_results[sample(nrow(ht_simulation_results), 100), ]
#write.csv(demo_mortality, "./output/demo_mortality.csv")
#write.csv(demo_simulation, "./output/demo_simulation.csv")

```


## 2.4 Useage

### 2.4.1 Interactive

Create interactive plot for decision making.

```{r plotly, echo=FALSE, eval = FALSE}

decision_point <- mortality_sample_plot +
  geom_point(data = mortality_data, 
             aes(x = sample_size, y = n_mortality, 
                 text = paste("Sample size:", sample_size, "<br>", 
                             survival_formatted),  color = decision),
             alpha = 0, size = 1)+
  scale_color_manual(
    values = c(
      "Accept H0" = "darkgreen", 
      "Reject H0" = "darkred", 
      "Continue" = "grey50"
    )
  )

decision_point$layers <- decision_point$layers[
  !sapply(decision_point$layers, function(x) inherits(x$geom, "GeomText"))
]

decision_point_plotly <- ggplotly(decision_point, tooltip = "text")
tmp <- tempfile(fileext = ".html")
htmlwidgets::saveWidget(decision_point_plotly, tmp)
browseURL(tmp)

```


### 2.4.2 Lookup mortality

```{r lookup_mort, eval = FALSE}
# lookup specific rows (sample, mortality)
lookup_mortality_data(mortality_data, 400, 7) 
lookup_mortality_data(mortality_data, 360, 7) 

# need to add lookup for simulation results too, so we can quickly identify decision points
```

### 2.4.3 Sequential decision

```{r seq_decision, eval = FALSE}

# Plot a specific simulation scenario when working live
# e.g.,  assumed true survival is 95% at interval 300, what do sequential test routes to rejecting null hypothesis this look like? 
# At 300 fish 
# We have a cumulative re-catch of 278 (22 missing fish)
# We have observed 7 dead fish 

# Survival probability of recaptured fish
#271/278 = 97.48%
# Recapture rate
#278/300 = 92.7%
# These show which lookup conditions to input for the simulation

#We then measure mortality with 3 methods and perform sensitivity analysis
#Run with defaults first, then can input the assumed_true_survival (0.98) and recapture_rate_scenario (1) afterwards

my_decisions <- lookup_seq_95(
  exposed_fish = 300,
  recaptured_fish = 270, 
  observed_deaths = 4,
  mortality_table = mortality_data,
  cumulative_data = cumulative_all
)

#needs to 
# show stopping proportions plot
# show simulated trials plot
# show sensitivity analysis plot

```


## 2.5 Visualise trial outcomes

Using the simulation output, we can now determine what scenarios are most likely to occur during sequential testing. This allows us to determine our stopping rate at given sample itnervals and given mortality estimates.

```{r simulation_plot, eval=FALSE}

# Plot simulations for 90 - 100% survival, 90% and 100% recapture using the neutral method. Mortality in missing fish assumed to be random (like observed fish)

journey_plot_recapture <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c(0.90, 0.91, 0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 0.98, 0.99,  1),
    recapture_rates_to_show = c(0.9, 1.0),
  decision_methods_to_show = c("neutral"),
  n_journeys = 1000
)

journey_plot_recapture$layers <- journey_plot_recapture$layers[
  !sapply(journey_plot_recapture$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recapture

```


```{r simulation_plot_examples, eval=FALSE}

journey_plot_recapture <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c(0.90, 0.95),
    recapture_rates_to_show = c(1.0),
  decision_methods_to_show = c("neutral"),
  n_journeys = 1000
)

journey_plot_recapture$layers <- journey_plot_recapture$layers[
  !sapply(journey_plot_recapture$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recapture <- journey_plot_recapture +
  labs(title = NULL, subtitle = NULL)

ggsave(filename="journey_plot_recapture.svg", plot=journey_plot_recapture,
       device = "svg",units="cm", width=24,height=16, dpi=300)

journey_plot_recapture1 <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c(0.97, 0.98),
    recapture_rates_to_show = c(1.0),
  decision_methods_to_show = c("neutral"),
  n_journeys = 1000
)

journey_plot_recapture1$layers <- journey_plot_recapture1$layers[
  !sapply(journey_plot_recapture1$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recapture1 <- journey_plot_recapture1 +
  labs(title = NULL, subtitle = NULL)

ggsave(filename="journey_plot_recapture1.svg", plot=journey_plot_recapture1,
       device = "svg",units="cm", width=24,height=16, dpi=300)


journey_plot_recapture2 <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c(0.99, 1),
    recapture_rates_to_show = c(1.0),
  decision_methods_to_show = c("neutral"),
  n_journeys = 1000
)

journey_plot_recapture2$layers <- journey_plot_recapture2$layers[
  !sapply(journey_plot_recapture2$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recapture2 <- journey_plot_recapture2 +
  labs(title = NULL, subtitle = NULL)

journey_plot_recapture2

ggsave(filename="journey_plot_recapture2.svg", plot=journey_plot_recapture2,
       device = "svg",units="cm", width=24,height=16, dpi=300)

# Plot simulations for specific survival and recapture rate, three mortality estimates

journey_plot_recap_method <- create_journey_plot_ht_comprehensive(
  mortality_data,
  ht_simulation_results_fixed,
  survival_rates_to_show = c( 0.98),
    recapture_rates_to_show = c(0.90),
  decision_methods_to_show = c("neutral", "aggressive", "conservative"),
  n_journeys = 1000
)

journey_plot_recap_method$layers <- journey_plot_recap_method$layers[
  !sapply(journey_plot_recap_method$layers, function(x) inherits(x$geom, "GeomText"))
]

journey_plot_recap_method

```

Need to add ability to plot specific simulations e.g.,

We're at interval 300
- what do scenarios which end here look like?
- what do scenarios which end at the next interval look like?

## 2.6 Stopping decisions

Now, we need to understand when we are most likely to stop, at a given sample interval, for a given true survival rate and recapture 

```{r create_cumulative}

ht_neutral <- ht_simulation_results %>% 
  filter(decision_method == "neutral")

ht_aggressive<- ht_simulation_results %>% 
  filter(decision_method == "aggressive")

ht_conservative <- ht_simulation_results %>% 
  filter(decision_method == "conservative")

cumulative_data_neutral <- cumulative_stopping(ht_neutral)
cumulative_data_conservative<- cumulative_stopping(ht_conservative)
cumulative_data_aggressive<- cumulative_stopping(ht_aggressive)

# Combine and add method labels to create complete lookup table for stopping decisions
cumulative_all <- bind_rows(
  cumulative_data_aggressive %>% mutate(decision_method = "aggressive"),
  cumulative_data_neutral %>% mutate(decision_method = "neutral"),
  cumulative_data_conservative %>% mutate(decision_method = "conservative")
)

#write.csv(cumulative_all, "./output/cumulative_all.csv")

```

```{r stopping_example, eval=FALSE}

#use lookup table to find specific example of stopping point for a given scenario 
# How many have stopped and accepted, rejected or continued by now using a given method?
# Lookup from cumulative table

stopping_93_recap_all_methods <- cumulative_all %>%
  filter(
    true_survival_rate == 0.98,
    recapture_rate == 0.93
  ) %>%
  select(
    decision_method,
    interval,
    prop_accept_by_now,
    prop_reject_by_now,
    prop_continue,
    prop_stopped_by_now
  ) %>%
  arrange(decision_method, interval)

#Alternatively, we can look just at how many trials have stopped using a specific method
# Given survival, recapture rate, how many trials have stopped by a given sample interval?
stopping_93_recap_method <- cumulative_all %>%
  filter(
    true_survival_rate == 0.98,
    recapture_rate == 1
  ) %>%
  select(interval, decision_method, prop_stopped_by_now)

# Pivot to wide format for easy comparison
stopping_93_recap_method <- stopping_93_recap_method %>%
  pivot_wider(names_from = decision_method, values_from = prop_stopped_by_now)

```


### 2.6.1 Inconclusivity rates

Assuming a max interval size of 1000

```{r inconclusivity_by_method, eval=FALSE}

inconclusivity_by_method <- cumulative_all %>%
  filter(interval == 1000) %>%
  select(true_survival_rate, recapture_rate, decision_method, prop_continue) %>%
  pivot_wider(names_from = recapture_rate, 
              values_from = prop_continue,
              names_prefix = "recapture_") %>%
  mutate(across(starts_with("recapture_"), ~ round(.x * 100, 1))) %>%
  arrange(decision_method, true_survival_rate)

inconclusivity_by_method

```

### 2.6.2 Visualise stopping decisions

We can now visualise our stopping decisions to provide a visual reference for when continuing sampling is a worthy trade-off

```{r plot_cumulative_stopping, eval=FALSE}

# Plot all survival and recapture conditions as a facet matrix 

plot_cumulative_stopping(cumulative_all, 
                         decision_method_filter = "neutral")

#decision_method_filter = "conservative"
#decision_method_filter = "aggressive"

```

```{r eval=FALSE}

#Plot specific scenario showing difference across recapture method
stopping_plot2 <- plot_cumulative_stopping(cumulative_all,
                                          survival_filter = 0.98, 
                                          recapture_filter = 0.9)
stopping_plot2


ggsave(filename="stopping_plot2.svg", plot=stopping_plot2,
       device = "svg",units="cm", width=24,height=10, dpi=300)

#Plot needs to be arranged correctly

```

Make sure all our plot methods arrange the recapture method as aggressive, netural, conservative

## 2.7 Visualise Type I error

We can examine the influence of our recapture estimate on type I error.

```{r cumulative_plot, eval=FALSE}

#Examine acceptance rate across all

cumulative_plot <- ggplot(cumulative_all, 
                         aes(x = interval, y = prop_accept_by_now, 
                             color = decision_method, linetype = decision_method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("aggressive" = "blue", "neutral" = "black", "conservative" = "red")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = c(100, 200, 300, 400, 500, 600, 700, 800, 900, 1000)) +
  labs(
    title = "Cumulative Acceptance Rate by Sample Interval and Decision Method",
    x = "Sample Interval",
    y = "Cumulative Proportion Accepting H0",
    color = "Decision Method",
    linetype = "Decision Method"
  ) +
  theme_JN() +
  facet_grid(recapture_rate ~ true_survival_rate, 
             labeller = labeller(recapture_rate = function(x) paste("Recapture:", x),
                               true_survival_rate = function(x) paste("Survival:", x)))

cumulative_plot


```

```{r cumulative_plot_9398, eval=FALSE}

#Examine acceptance rate across all methods for specific survival and recapture

cumulative_plot_9398 <- cumulative_all %>% 
  filter(true_survival_rate == 0.98, recapture_rate == 0.90) %>%
  select(decision_method, interval, prop_accept_by_now, prop_reject_by_now, prop_continue) %>%
  pivot_longer(cols = c(prop_accept_by_now, prop_reject_by_now, prop_continue),
               names_to = "outcome", 
               values_to = "proportion") %>%
  mutate(
    outcome = case_when(
      outcome == "prop_accept_by_now" ~ "Accept H0",
      outcome == "prop_reject_by_now" ~ "Reject H0", 
      outcome == "prop_continue" ~ "Continue"
    )
  ) %>%
  ggplot(aes(x = interval, y = proportion, 
             color = decision_method, linetype = decision_method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("aggressive" = "blue", "neutral" = "black", "conservative" = "red")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = c(100, 200, 300, 400, 500, 600, 700, 800, 900, 1000)) +
  labs(
    title = "Cumulative Trial Outcomes by Sample Interval (98% Survival, 93% Recapture)",
    x = "Sample Interval",
    y = "Cumulative Proportion",
    color = "Decision Method",
    linetype = "Decision Method"
  ) +
  theme_JN() +
  theme(legend.position = "bottom") +
  facet_wrap(~outcome, scales = "free_y") +
  coord_cartesian(clip = "off")

cumulative_plot_9398


ggsave(filename="cumulative_plot_9398.svg", plot=cumulative_plot_9398,
       device = "svg",units="cm", width=24,height=10, dpi=300)


```


We may then consider two things (1) Where the 'grey zone' exists when we reject null hypothesis and it may actually be true (e.g., the level of acceptance in 0.94 - 0.97). This is our type I error rate, which increase when trying to estimate close to thresholds.
(2) how we can control type I error if we use neutral method as it influences acceptance the least.

```{r sensitivity_plot, eval=FALSE}

ht_summary <- create_stopping_summary_ht(ht_simulation_results)

# Show how decision method affects outcomes across recapture rates
sensitivity_data <- ht_summary$trial_overview %>%
  select(true_survival_rate, recapture_rate, decision_method, prop_accept, prop_continue)

sens_plot <- ggplot(sensitivity_data, aes(x = recapture_rate, y = prop_accept, 
                            color = decision_method, linetype = decision_method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("aggressive" = "blue", "neutral" = "black", "conservative" = "red")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Acceptance Rate by Recapture Rate and Decision Method (98% survival)",
    x = "Recapture Rate",
    y = "Proportion Accepting H0",
    color = "Decision Method",
    linetype = "Decision Method"
  ) +
  theme_JN() +
  facet_wrap(~true_survival_rate)

sens_plot

```


```{r eval=FALSE}

# For a specific example e.g., 98% true survival

sens_plot2 <- ggplot(sensitivity_data %>%
                       filter(true_survival_rate == 0.98), aes(x = recapture_rate, y = prop_accept, 
                            color = decision_method, linetype = decision_method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("aggressive" = "blue", "neutral" = "black", "conservative" = "red")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Acceptance Rate by Recapture Rate and Decision Method (98% survival)",
    x = "Recapture Rate",
    y = "Proportion Accepting",
    color = "Decision Method",
    linetype = "Decision Method"
  ) +
  theme_JN() +
  coord_cartesian(clip = "off")

sens_plot2

ggsave(filename="sens_plot2.svg", plot=sens_plot2,
       device = "svg",units="cm", width=16,height=10, dpi=300)


# We can see that the effect is worst when recapture rate is lower - it reflects our real world confidence!

```


#Need function for finding minimum sample size for given proportion and confidence interval estimate 

# 3. Hypothese testing

## 3.1 Compliance testing

Scenario: Observed 100% survival
          Observed 99% survival
          Observed 97% survival
          Observed 96% survival
          
Objective: Does pump meet regulatory threshold (≥98%)?

An Exact Binomial test can be applied to determine if an observed survival rate is significantly higher (greater) or lower (less) than a given compliance threshold (p = 0.98).The test can be applied where observed survival is 99% or 100% and statistical evidence to support the survival estimate is desired without a control group. When survival is marginally compliant, or non-compliant, the test may be applied but statistical comparison to a control group is preferred in these scenarios.

Excessively large sample sizes will be required to show 99% is > 98%, and 97% < 98%. This is unethical, and should therefore use confidence interval estimation and/or inferiority (impact survival < control survival) or superiority (impact mortality > control mortality) testing.

### 3.1.1 Test

```{r compliance_test, eval= FALSE}

#Superiority approach

# Example data: 100% survival at minimum sample size of 188 where 95% CI = 98.4% - 100%.
binom.test(x = 188, n = 188, p = 0.98, alternative = "greater")
#p = 0.02241

# Example data: 99% survival at minimum sample size of 200 where 95% CI = 96.43% - 99.73%.
binom.test(x = 198, n = 200, p = 0.98, alternative = "greater")
#p = 0.2351
#See power analysis below for required sample size to show 99% exceeds 98% compliance

#Inferiority approach

# Example data: 96% survival at minimum sample size of 200 where 95% CI = 92.31% - 97.96%
binom.test(x = 192, n = 200, p = 0.98, alternative = "less")
#p = 0.04934

# Example data: 97% survival at minimum sample size of 600 where 95% CI = 95.31 - 98.09%
binom.test(x = 582, n = 600, p = 0.98, alternative = "less")
#p = 0.06104
#See power analysis below for required sample size to show 97% does not meet 98% compliance

# check what this power measurement is doing and decide whether to include it
# Power when true survival = 99%
# power <- 1 - pbinom(957, 969, 0.99) 
# power

```

### 3.1.2 Power

```{r compliance_power, eval = FALSE}

result_greater <- find_compliance_sample_size(
  true_rate = 0.99, 
  test_direction = "greater", 
  target_power = 0.8
)
print(result_greater)
# Sample size of at least 969 required to observe an exact 99% survival that statistically exceeds 98%
# Reject (compliance ≤  0.98 ) if ≥  957 / 969  survivors
# Example data: 99% survival at minimum sample size of 200 where 95% CI = 96.43% - 99.73%.
# binom.test(x = 957, n = 969, p = 0.98, alternative = "greater")
# p = 0.04973

result_less <- find_compliance_sample_size(
  true_rate = 0.97, 
  test_direction = "less", 
  target_power = 0.8
)
print(result_less)
# Sample size of at least 1469 required to observe an exact 97% survival that statistically does not meet 98%
# Reject (compliance ≥  0.98 ) if ≤  1430 / 1469  survivors
# Example data: 97% survival at minimum sample size of 600 where 95% CI = 95.31 - 98.09%
# binom.test(x = 1430, n = 1469, p = 0.98, alternative = "less")
# p = 0.04931

```


What about when we observe 98% survival?

Exact binomial test for compliance cannot be applied when observed survival = 98%. Instead, we rely on CI estimation, and can also make interpretation from comparisons with a control group.

Group comparison test directional considerations
  - Two-sided: Is there any statistical difference between pump and control?
  - One-sided (less): Is pump survival inferior to control survival?
      matrix(c(impact_surv, impact_died, control_surv, control_died), nrow=2, byrow=T)
      fisher.test(matrix, alternative = "less")
  - One-sided (greater): Is control survival superior to impact survival?
      matrix(c(control_surv, control_died, impact_surv, impact_died), nrow=2, byrow=T)
      fisher.test(matrix, alternative = "greater")
  - Use a survival-first matrix. It is critical the first row of the matrix represents the direction of compliance
  e.g., impact first = alternative tests against control
  
For consistency, there is no need to consider reversing the matrix and asking if control is superior. We should always determine whether pump survival is inferior. If the two observed proportions are close (e.g., 94% impact, 95% control), we can switch to a TOST.
  
To examine the question of 'Is pump acceptable despite meeting threshold exactly?', we can take a difference or inferiority test approach. The same approach will also be followed for determining test choice where we ask 'Does the pump cause harm?'

## 3.2 Difference testing

### 3.2.1 Test

- Two-sided: Is there any statistical difference between pump and control?

Scenario: Observed 98% survival
          Observed ≤ 97% survival

Objective: Is there any difference between pump and control?

A two-sided Fisher's Exact test can be applied to determine if an observed survival rate is significantly different (non-directional) than a control survival. The test can be applied where observed survival is ≤ 98% and statistical evidence to support the survival estimate is desired with a control group.

Excessively large sample sizes will be required to show 98% is different to 100%. This is unethical, and should therefore use confidence interval estimation primarily as little is gained by showing difference with a control group.

```{r difference_test, eval=FALSE}

# Example data 98% survival at minimum sample size of 400 where 95% CI = 96.11% - 98.98%
observed_impact_surv = 0.98
impact_n = 400
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 134 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.219
#See power analysis below for required sample size to show a difference between impact (98%) and control (100%) exists


# Example data: 97% survival at minimum sample size of 600 where 95% CI = 95.31 - 98.09%
observed_impact_surv = 0.97
impact_n = 600
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 200 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.01022
# Suggests sample size could actually be lower than if using CI alone
#See power analysis below for required sample size to show a difference between impact (97%) and control (100%) exists


# Example data: 90% survival at minimum sample size of 100 where 95% CI = 82.56 -  94.48%
observed_impact_surv = 0.97
impact_n = 100
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 33 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.5738
#See power analysis below for required sample size to show a difference between impact (90%) and control (100%) exists


```

### 3.2.1 Power

```{r difference_power, eval = FALSE}

# Test for any difference between groups when impact survival = 98%
find_min_sample_size_fisher(p_impact = 0.98, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "two.sided")

# sample size of at least impact n = 735, control n = 245 to  show a difference between impact (98%) and control (100%) exists

# observed_impact_surv = 0.98
# impact_n = 735
# surv_impact <- observed_impact_surv*impact_n
# dead_impact <- impact_n - surv_impact

# observed_control_surv = 1
# control_n = 245 #approx 1:3
# surv_control <- observed_control_surv*control_n
# dead_control <- control_n - surv_control

# survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
#                              nrow = 2, byrow = TRUE)
# fisher.test(survival_matrix, alternative = "two.sided")
#p = 0.02944

# Test for any difference between groups when impact survival = 97%
find_min_sample_size_fisher(p_impact = 0.97, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "two.sided")

# sample size of at least impact n = 492, control n = 164 to  show a difference between impact (97%) and control (100%) exists

 # observed_impact_surv = 0.97
 # impact_n = 492
 # surv_impact <- observed_impact_surv*impact_n
 # dead_impact <- impact_n - surv_impact
 # 
 # observed_control_surv = 1
 # control_n = 164 #approx 1:3
 # surv_control <- observed_control_surv*control_n
 # dead_control <- control_n - surv_control
 # 
 # survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
 #                              nrow = 2, byrow = TRUE)
 # fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.02885

# Test for any difference between groups when impact survival = 90%
find_min_sample_size_fisher(p_impact = 0.9, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "two.sided")

# sample size of at least impact n = 147, control n = 49 to  show a difference between impact (90%) and control (100%) exists

 # observed_impact_surv = 0.9
 # impact_n = 147
 # surv_impact <- observed_impact_surv*impact_n
 # dead_impact <- impact_n - surv_impact
 # 
 # observed_control_surv = 1
 # control_n = 49 #approx 1:3
 # surv_control <- observed_control_surv*control_n
 # dead_control <- control_n - surv_control
 # 
 # survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
 #                              nrow = 2, byrow = TRUE)
 # fisher.test(survival_matrix, alternative = "two.sided")

#p = 0.0247


```

Powering the test to detect a difference is not statistically meaningful when we observe marginally compliant survival, or where control survival is imperfect and we need to understand if the pump was actually harmful. 

## 3.3 Inferiority testing

### 3.3.1 Test

- One-sided: Is impact survival statistically lower than control survival?

Scenario: Observed 98% survival
          Observed ≤ 97% survival

Objective: Is pump survival inferior to control survival? Does the pump cause harm?

A one-sided Fisher's Exact test can be applied to determine if an observed impact survival rate is significantly lower (less) than a control survival. The test can be applied where observed survival is ≤ 98% and statistical evidence to support the survival estimate is desired with a control group.


```{r inferiority_test, eval = FALSE}

# Example data 98% survival at minimum sample size of 400 where 95% CI = 96.11% - 98.98%
observed_impact_surv = 0.98
impact_n = 400
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 134 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "less")

# p = 0.09737
#See power analysis below for required sample size to show impact (98%) survival is less than control (100%).


# Example data: 97% survival at minimum sample size of 600 where 95% CI = 95.31 - 98.09%
observed_impact_surv = 0.97
impact_n = 600
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 200 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "less")

#p = 0.005284
# Suggests sample size could actually be lower than if using CI alone
#See power analysis below for required sample size to show a difference between impact (97%) and control (100%) exists


# Example data: 90% survival at minimum sample size of 100 where 95% CI = 82.56 -  94.48%
observed_impact_surv = 0.97
impact_n = 100
surv_impact <- observed_impact_surv*impact_n
dead_impact <- impact_n - surv_impact

observed_control_surv = 1
control_n = 33 #approx 1:3
surv_control <- observed_control_surv*control_n
dead_control <- control_n - surv_control

survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control), 
                             nrow = 2, byrow = TRUE)
fisher.test(survival_matrix, alternative = "less")

#p = 0.4219
#See power analysis below for required sample size to show a difference between impact (90%) and control (100%) exists


```


The key observation is that we increase statistical power when using a one-sided test. We are able to justify this as the pump is never likely to exceed control survival, nor is significance in this direction of interest.

### 3.2.1 Power

```{r inferiority_power, eval = FALSE}

# Test for any difference between groups when impact survival = 98%
find_min_sample_size_fisher(p_impact = 0.98, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "less")

# sample size of at least impact n = 681, control n = 227 to show impact (98%) is inferior to control (100%)

 # observed_impact_surv = 0.98
 # impact_n = 681
 # surv_impact <- observed_impact_surv*impact_n
 # dead_impact <- impact_n - surv_impact
 # 
 # observed_control_surv = 1
 # control_n = 227 #approx 1:3
 # surv_control <- observed_control_surv*control_n
 # dead_control <- control_n - surv_control
 # 
 # survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
 #                              nrow = 2, byrow = TRUE)
 # fisher.test(survival_matrix, alternative = "less")
#p = 0.01723

# Test for any difference between groups when impact survival = 97%
find_min_sample_size_fisher(p_impact = 0.97, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "less")

# sample size of at least impact n = 453, control n = 151 to show impact (97%) is inferior to control (100%) exists

 # observed_impact_surv = 0.97
 # impact_n = 453
 # surv_impact <- observed_impact_surv*impact_n
 # dead_impact <- impact_n - surv_impact
 # 
 # observed_control_surv = 1
 # control_n = 151 #approx 1:3
 # surv_control <- observed_control_surv*control_n
 # dead_control <- control_n - surv_control
 # 
 # survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
 #                              nrow = 2, byrow = TRUE)
 # fisher.test(survival_matrix, alternative = "less")

#p = 0.01693

# Test for any difference between groups when impact survival = 90%
find_min_sample_size_fisher(p_impact = 0.9, p_control = 1,
                                    alpha = 0.05, power_target = 0.8, 
                                    group_ratio = 3, step_size = 1, 
                                    alternative = "less")

# sample size of at least impact n = 135, control n = 45 to show impact (90%) is inferior to control (100%) exists

 observed_impact_surv = 0.9
 impact_n = 135
 surv_impact <- observed_impact_surv*impact_n
 dead_impact <- impact_n - surv_impact

 observed_control_surv = 1
 control_n = 45 #approx 1:3
 surv_control <- observed_control_surv*control_n
 dead_control <- control_n - surv_control

 survival_matrix <- matrix(c(surv_impact, dead_impact, surv_control, dead_control),
                              nrow = 2, byrow = TRUE)
 fisher.test(survival_matrix, alternative = "less")

#p = 0.01532


```

What about when we observe non-compliant pump survival but control survival was also < 100%?

We can frame the question in two approaches:

Inferiority test: Does the pump harm fish more than background mortality (e.g., mortality caused by net or handling control)? This approaches uses the existing one-sided Fisher's exact test and is defensible when impact mortality is  lower than control mortality (e.g., it causes an additional mortality to background). However, it can theoretically detect any difference, and a significant result could therefore falsely conclude pump is harmful if there is enough statistical power to detect a small difference (e.g., 1%). Prove the pump is worse for fish that enter the pump than fish which do not enter the pump. Pump causes demonstrable harm.

Equivalence test: Is pump performance within 2% of background mortality (our new accepted standard based on control background mortality)? This approaches uses TOST ...... 

## 3.5 Equivelance testing

### 3.5.1 Test

TOST is unlikely to be suitable here because it is a very conservative approach. The test requires the entire confidence interval to fit within the given equivalence bounds. Therefore, sample sizes are likely to be very large and an even higher demand than difference or inferiority testing.

```{r tost_test, eval= FALSE}

# Example data: 90% pump survival and 91% control survival
# Test for equivalence
TOSTtwo.prop(prop1 = 0.9,      # e.g., 0.94
             prop2 = 0.91,   # e.g., 0.96  
             n1 = 800,                 # e.g., 300
             n2 = 800,              # e.g., 200
             low_eqbound = -0.02,             # -2% margin
             high_eqbound = 0.02,             # +2% margin
             alpha = 0.05)

```

### 3.5.1 Power

```{r tost_power, eval = FALSE}

#Not working

# Power analysis
powerTOSTtwo.prop(alpha = 0.05,
                  statistical_power = 0.8,
                  prop1 = 0.9,              # expected impact
                  prop2 = 0.91,              # expected control
                  N = NULL,                  # calculate required N
                  low_eqbound_prop = -0.02,  # -2% margin
                  high_eqbound_prop = 0.02)  # +2% margin


```


Overall, our entire fish survival framework comes down to four considerations.

- 95% Confidence interval estimation
  Always applied to ensure precise survival estimate. Used exclusively for high survival scenarios to prevent unethical sample designs (98%, 99%, 100%)
  
- Exact Binomial compliance test
  May be applied to show compliance if a control group is committed, or when survival is high and desire to     statistically prove > 98%.
  
- Fisher's Exact test (two.sided difference)
  May be applied to show a difference between impact and control fish when survival < 99%. Not advised to apply as we are interested in pump harm.
  
- Fisher's Exact test (one.sided inferiority)
  The primary test consideration when impact survival < 98%. Show inferiority from impact group to control group.
  
 Next to-do:
 
 - Build up the visualizations to show these choices e.g., power curves.
 - Sequential testing procedure when using one-sided fisher's exact. What are the lowest possible sample sizes, and create a stopping decision criteria like the mortality table Account for alpha spend, data peeking? Not sure.
 
 
# 4. Group sequential design

## 4.1 

```{r}

# One-sided, 4 interim analyses, O'Brien-Fleming boundaries
design <- gsDesign(
  k = 10,                # Number of interim analyses
  test.type = 1,        # 1 = one-sided (H1: p_impact < p_control)
  alpha = 0.05,         # Type I error
  beta = 0.20,          # Type II error (Power = 80%)
  sfu = "OF"            # O'Brien-Fleming spending function
)

# Print stopping boundaries (upper bounds are for efficacy)
print(design)

```
